Approximately Exact Calculations for Linear Mixed Models

### 2015-09-24 ejs.tex
ejs version

\documentclass{report}

\usepackage{amsmath,amsfonts}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{caption,subcaption}
%\usepackage{color}
%\usepackage{etoolbox}
\usepackage{float}
\usepackage{fullpage}
%\usepackage{mathtools}
\usepackage{natbib}
\usepackage[inline]{enumitem}
\usepackage{subdepth}
%\usepackage{appendix}
%\usepackage{setspace}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}

\renewcommand\thesection{\arabic{section}}
\usepackage{xspace}
\newcommand{\textcompute}{\textsf}
\newcommand{\N}{\text{N}} % Normal 
\newcommand{\R}{\textcompute{R}\xspace}
\newcommand{\IG}{\text{InvGam}}
\newcommand{\RL}{f}
\newcommand{\RLorig}{\text{RL}}
\newcommand{\logRL}{\log\RL}
\newcommand{\logRLorig}{\log\RLorig}
\newcommand{\sigssq}{\sigma_s^2}
\newcommand{\sigesq}{\sigma_e^2}
\newcommand{\sshat}{\hat\sigma^2_e,\hat\sigma^2_s}
\newcommand{\logRLss}{\logRL(\sigesq,\sigssq)}
\newcommand{\logRLssorig}{\logRLorig(\sigesq,\sigssq)}
\newcommand{\ass}{a_j\sigssq + \sigesq}
\newcommand{\abss}{a_j\sigssq + b_j\sigesq}
\newcommand{\fss}{f(\sigesq,\sigssq)\,}
%\newcommand{\mrle}{MRLE}
\newcommand{\mrle}{$\argsup\log f$}
%\newcommand{\lone}{\textbf{l}_1}
%\newcommand{\pone}{\textbf{p}_1}
%\newcommand{\ltwo}{\textbf{l}_2}
%\newcommand{\ptwo}{\textbf{p}_2}
%\newcommand{\pthree}{\textbf{p}_3}
%\newcommand{\given}{\,|\,}
\newcommand{\g}{\,|\,}
\newcommand{\maxit}{\textcompute{maxit}}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argsup}{\operatornamewithlimits{argsup}}
\newcommand*\DNA{\textsc{dna}}
\newcommand*\Let[2]{\State #1 $\gets$ #2}
\renewcommand{\bibname}{References}
\linespread{1.6}

@@

\title{Approximately Exact Calculations for Linear Mixed Models}
\author[1]{Michael Lavine}
\author[2]{Andrew Bray}
\author[3]{James Hodges}
\affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
\affil[2]{Reed College, Portland, OR, 97202 USA}
\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN, 55455 USA}

\maketitle

@@

@@

\section{Examples}
\label{sec:examp}
Section~\ref{sec:examp}'s examples use $\epsilon = 1$ and $M = 7$.
Computations were performed on an iMac that was new in 2011, having a 2.7 GHz Intel Core i5 processor and 4GB of memory.
\subsection{HMO premiums}
\subsubsection{Introduction to the Data}
Our first example is a traditional linear mixed model previously analyzed in \citet{hodges:98},  \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014}. \cite{wakefield:1998} reported a bimodal log posterior density for $(\sigesq,\sigssq)$.  Quoting from \cite{henn&hodges:2014},
\begin{quote}
\itshape
\dots the HMO data set describes 341 HMOs [Health Maintenance Organizations] located in 45 states or similar political jurisdictions.  Each jurisdiction had between 1 and 31 plans with a median of 5 plans.  The data set originally was analysed to assess the cost of moving military retirees and dependents from a Department of Defense health plan to plans serving the US civil service.  \dots.  Specifically, the model is
\begin{equation*}
\begin{split}
	y_{ij} &= \alpha_i + \epsilon_{ij}\\
	\alpha_i &= \varrho_0 + \varrho_1x_{1i} + \varrho_2x_{2i} + \zeta_i,
\end{split}
\end{equation*}
where the fixed effects in $\alpha_i$ include an intercept, jurisdiction-average hospital expenses per admission ($x_{1i}$) and an indicator for plans in New England states ($x_{2i}$).

@@

Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b show a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

\subsubsection{A Bayesian Analysis}
 \cite{hodges:98}, \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014} report  Bayesian analyses of the HMO data.  Here we reproduce the analysis from \cite{hodges:98} which used inverse Gamma priors for $(\sigesq,\sigssq)$ with $\alpha_e = 1$; $\beta_e = 0$; $\alpha_s = 1.1$; and $\beta_s =0.1$.  (We don't defend the prior; we use it so we can compare to Hodges.)

@@

\subsection{Global Mean Surface Temperature}
We reanalyze a data set in \cite{hodges:2013}, global mean surface temperatures (GMST) from 1881 through 2005, depicted in Figure~\ref{fig:gmst-scatter}.  
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/gmst-scatter.pdf}
	\caption{Global mean surface temperature annually from 1881.
		The $y$-axis shows deviations from the overall mean in units of
		.01 degrees C.}
	\label{fig:gmst-scatter}
\end{figure}
As shown in \cite{ruppert_etal:2003}, many splines can be written as linear mixed models.  \cite{hodges:2013} fit a piecewise quadratic spline to the GMST data, though a piecewise cubic spline would look similar.  Both splines can be formulated as linear mixed models.  We follow his lead in fitting a quadratic spline with knots at 1880, 1884, 1888, \dots, 2004.  $X$ has three columns: $\textcompute{1}, \textcompute{year}, \textcompute{year}^2$.  $Z$ is $125 \times 30$: one row for each year; one column for each knot.
%\begin{equation*}
%Z =	\begin{bmatrix}
%		0 & 0 & \dots & 0\\
%		0 & 0 & \dots & 0\\
%		0 & 0 & \dots & 0\\
%		0 & 0 & \dots & 0\\
%		1 & 0 & \dots & 0\\
%		4 & 0 & \dots & 0\\
%		9 & 0 & \dots & 0\\
%		16 & 0 & \dots & 0\\
%		25 & 1 & \dots & 0\\
%		\vdots & \vdots & \vdots & \vdots\\
%		13924 & 12996 & \dots & 4\\
%		14161 & 13225 & \dots & 9\\
%		14400 & 13456 & \dots & 16\\
%		14641 & 13689 & \dots & 25\\
%	\end{bmatrix}; \hspace{.5cm} \Sigma_e = \mathbf{1}_{125}; \hspace{.5cm}
%		\Sigma_s = \mathbf{1}_{30}
%\end{equation*}
Because we fit a quadratic spline, the entries in $Z$ are squares.  $\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension; see \cite{hodges:2013} for details.  Following Hodges, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
\begin{equation*}
Z =	\begin{bmatrix} 
		0 & 0 & \dots & 0\\
		\vdots & \vdots & \vdots & \vdots\\
		10.97143 & 10.2521 & \dots & 0.01219048\\
		11.15505 & 10.42971 & \dots & 0.01904762\\
	\end{bmatrix}
\end{equation*}
Centering and scaling changes only the scale on which $\sigssq$ is measured; we do it to more easily compare our result to Hodges'.

@@

\item A \textcompute{box} $b$ is a list consisting of
	 \begin{itemize}
	 \item upper and lower limits on $\sigesq$,
	 \item upper and lower limits on $\sigssq$,
	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
	   or straddles each of the $s_z+2$ lines.
	 \end{itemize}
\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
	a box $b$ should be divided more finely.
\item \textcompute{splitbox} is a function that takes a box $b =
	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
	and returns the four boxes created by dividing each side of $b$ at its midpoint.
\end{enumerate}

\paragraph{\textcompute{findf}: The main function}
\begin{verbatim}
findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
                  ratio = FALSE, lognote = "summary") {

    inactive <- list()  # a list of inactive boxes
    ninact <- 0  # the number of inactive boxes

    active <- list(startbox)  # a list of active boxes
    nact <- 0  # the number of active boxes

    lowbound <- -Inf  # lowerbound on max(log(f))
    iter <- 0  # iteration number

    while (nact > 0 && iter < maxit) {
        # Find the lower bound of each box and the maximum of the lower bounds.  For
        # each active box, either make it inactive or divide it.
        low.act <- max(vapply(X = active, FUN = function(box) {
            box$bounds[1]
        }, FUN.VALUE = 0.1))
        lowbound <- max(lowbound, low.act)
        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
            # which boxes become inactive?
        nkill <- sum(kill)
        if (nkill > 0) {
            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
            # add those boxes to the inactive list
        }
        ninact <- length(inactive)
        kids <- list()  # subdivisions of active boxes
        nkids <- 0
        for (i in which(!kill)) {  # split active boxes into 4 parts
            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
            nkids <- nkids + 4
        }
        active <- kids
        nact <- length(active)

@@

\paragraph{\textcompute{killfunc}: determine whether to make a box inactive}
\begin{verbatim}
# conditions under which a box becomes inactive
killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
    # lb is a global (within startbox) lower bound on max(log(f));
    #      it changes at each iteration.
    #  M, eps, delE, delS stay constant throughout the iterations.
    cond.low <- box$bounds[2] < lb - M
    cond.eps <- diff(box$bounds) < eps
    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
                            diff(box$lims.sigsqe) < delE)
    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
                            diff(box$lims.sigsqs) < delS)
    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
}
\end{verbatim}

\paragraph{\textcompute{splitbox}: split a box into four children}
\begin{verbatim}
splitbox <- function(box, lines) {
    # split a box into four children
    NW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
        lines = lines))

    NE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
        lines = lines))

    SW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
        lines = lines))

    SE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
        lines = lines))

    return(list(NW, NE, SW, SE))
}

\end{verbatim}

>>>>

%\documentclass{report}
\documentclass[ejs]{imsart}

\usepackage{amsmath,amsfonts}
\usepackage{graphicx}
\usepackage{caption,subcaption}
\usepackage{float}
\usepackage{fullpage}
\usepackage[numbers]{natbib}
\usepackage[inline]{enumitem}
\usepackage{subdepth}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{xspace}

\startlocaldefs
\renewcommand\thesection{\arabic{section}}
\newcommand{\textcompute}{\textsf}
\newcommand{\N}{\text{N}} % Normal 
\newcommand{\R}{\textcompute{R}\xspace}
\newcommand{\IG}{\text{InvGam}}
\newcommand{\RL}{f}
\newcommand{\RLorig}{\text{RL}}
\newcommand{\logRL}{\log\RL}
\newcommand{\logRLorig}{\log\RLorig}
\newcommand{\sigssq}{\sigma_s^2}
\newcommand{\sigesq}{\sigma_e^2}
\newcommand{\sshat}{\hat\sigma^2_e,\hat\sigma^2_s}
\newcommand{\logRLss}{\logRL(\sigesq,\sigssq)}
\newcommand{\logRLssorig}{\logRLorig(\sigesq,\sigssq)}
\newcommand{\ass}{a_j\sigssq + \sigesq}
\newcommand{\abss}{a_j\sigssq + b_j\sigesq}
\newcommand{\fss}{f(\sigesq,\sigssq)\,}
\newcommand{\mrle}{$\argsup\log f$}
\newcommand{\g}{\,|\,}
\newcommand{\maxit}{\textcompute{maxit}}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argsup}{\operatornamewithlimits{argsup}}
\newcommand*\DNA{\textsc{dna}}
\newcommand*\Let[2]{\State #1 $\gets$ #2}
\renewcommand{\bibname}{References}
\endlocaldefs

\linespread{1.6}

@@

\begin{frontmatter}

\title{Approximately Exact Calculations for Linear Mixed Models}
\runtitle{Approximately Exact Calculations}

\begin{aug}

\author{ \fnms{Michael} \snm{Lavine}  \ead[label=e1]{lavine@math.umass.edu} }
\address{Department of Mathematics and Statistics\\ UMass Amherst\\ Amherst, MA 01003\\ USA}
\printead{e1}

\author{ \fnms{Andrew} \snm{Bray}  \ead[label=e2]{abray@reed.edu} }
\address{Reed College\\ 3203 Southeast Woodstock Boulevard\\ Portland, Oregon 97202-8199\\ USA}
\printead{e2}

\author{ \fnms{Jim} \snm{Hodges}  \ead[label=e3]{hodge003@umn.edu} }
\address{Division of Biostatistics\\
               School of Public Health\\
               University of Minnesota\\
               2221 University Ave SE, Suite 200\\
               Minneapolis, MN 55414\\
               USA
}
\printead{e3}

\runauthor{M. Lavine et al.}
\affiliation{Some University and Another University}

\end{aug}

@@

\begin{keyword}[class=MSC]
\kwd[Primary ]{62J05}
\kwd[; secondary ]{62F99}
\end{keyword}

\tableofcontents

\end{frontmatter}

@@

\section{Examples}
\label{sec:examp}
Section~\ref{sec:examp}'s examples use $\epsilon = 1$ and $M = 7$.
Computations were performed on an iMac that was new in 2011, having a 2.7 GHz Intel Core i5 processor and 4GB of memory.
\subsection{HMO premiums}
\subsubsection{Introduction to the Data}
Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98},  \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014}. \cite{wakefield:1998} reported a bimodal log posterior density for $(\sigesq,\sigssq)$.  Quoting from \cite{henn&hodges:2014},
\begin{quote}
\itshape
\dots the HMO data set describes 341 HMOs [Health Maintenance Organizations] located in 45 states or similar political jurisdictions.  Each jurisdiction had between 1 and 31 plans with a median of 5 plans.  The data set originally was analysed to assess the cost of moving military retirees and dependents from a Department of Defense health plan to plans serving the US civil service.  \dots.  Specifically, the model is
\begin{equation*}
\begin{split}
	y_{ij} &= \alpha_i + \epsilon_{ij}\\
	\alpha_i &= \varrho_0 + \varrho_1x_{1i} + \varrho_2x_{2i} + \zeta_i,
\end{split}
\end{equation*}
where the fixed effects in $\alpha_i$ include an intercept, jurisdiction-average hospital expenses per admission ($x_{1i}$) and an indicator for plans in New England states ($x_{2i}$).

@@

Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b show a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.  Further, about their Figure~2a, \cite{henn&hodges:2014} say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

\subsubsection{A Bayesian Analysis}
 \cite{hodges:98}, \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014} report  Bayesian analyses of the HMO data.  Here we reproduce the analysis from \cite{hodges:98} which used inverse Gamma priors for $(\sigesq,\sigssq)$ with $\alpha_e = 1$; $\beta_e = 0$; $\alpha_s = 1.1$; and $\beta_s =0.1$.  (We don't defend the prior; we use it so we can compare to \cite{hodges:98}.)

@@

\subsection{Global Mean Surface Temperature}
We reanalyze a data set in \cite{hodges:2013}, global mean surface temperatures (GMST) from 1881 through 2005, depicted in Figure~\ref{fig:gmst-scatter}.  
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/gmst-scatter.pdf}
	\caption{Global mean surface temperature annually from 1881.
		The $y$-axis shows deviations from the overall mean in units of
		.01 degrees C.}
	\label{fig:gmst-scatter}
\end{figure}
As shown in \cite{ruppert_etal:2003}, many splines can be written as linear mixed models.  \cite{hodges:2013} fit a piecewise quadratic spline to the GMST data, though a piecewise cubic spline would look similar.  Both splines can be formulated as linear mixed models.  We follow \cite{hodges:2013}'s lead in fitting a quadratic spline with knots at 1880, 1884, 1888, \dots, 2004.  $X$ has three columns: $\textcompute{1}, \textcompute{year}, \textcompute{year}^2$.  $Z$ is $125 \times 30$: one row for each year; one column for each knot.
Because we fit a quadratic spline, the entries in $Z$ are squares.  $\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension; see \cite{hodges:2013} for details.  Following \cite{hodges:2013}, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
\begin{equation*}
Z =	\begin{bmatrix} 
		0 & 0 & \dots & 0\\
		\vdots & \vdots & \vdots & \vdots\\
		10.97143 & 10.2521 & \dots & 0.01219048\\
		11.15505 & 10.42971 & \dots & 0.01904762\\
	\end{bmatrix}
\end{equation*}
Centering and scaling changes only the scale on which $\sigssq$ is measured; we do it to more easily compare our result to \cite{hodges:2013}'s.

@@

\item A \textcompute{box} $b$ is a list consisting of
	 \begin{itemize}
	 \item upper and lower limits on $\sigesq$,
	 \item upper and lower limits on $\sigssq$,
	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
	   or straddles each of the $s_z+2$ lines.
	 \end{itemize}
\item \textcompute{killfunc} is a function, defined within \textcompute{findf}, that determines whether
	a box $b$ should be divided more finely.
\item \textcompute{makebox} is a function that takes limits on $\sigesq$, $\sigssq$ and returns a
	\textcompute{box} structure. 
\item \textcompute{splitbox} is a function that takes a box $b =
	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
	and returns the four boxes created by dividing each side of $b$ at its midpoint.
\item \textcompute{getstatus} is a function that calculates whether a box is above, below, or straddles
	each line.
\item \textcompute{getbounds} is a function that computes $\{L^b_j, U^b_j\}$.
\item \textcompute{findlines} is a function that finds the lines for $j=1,\dots,s_z+1$ corresponding to $\logRLorig$.
\item \textcompute{addprior} is a function that adjusts the $s_z+1$ line and adds the $s_z+2$ line
	to account for a conjugate prior.
\end{enumerate}

%findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
%                  ratio = FALSE, lognote = "summary") {
%
%
%    inactive <- list()  # a list of inactive boxes
%    ninact <- 0  # the number of inactive boxes
%    
%    active <- list(startbox)  # a list of active boxes
%    nact <- 0  # the number of active boxes
%
%    lowbound <- -Inf  # lowerbound on max(log(f))
%    iter <- 0  # iteration number
%
%    while (nact > 0 && iter < maxit) {
%        # Find the lower bound of each box and the maximum of the lower bounds.  For
%        # each active box, either make it inactive or divide it.
%        low.act <- max(vapply(X = active, FUN = function(box) {
%            box$bounds[1]
%        }, FUN.VALUE = 0.1))
%        lowbound <- max(lowbound, low.act)
%        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
%            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
%            # which boxes become inactive?
%        nkill <- sum(kill)
%        if (nkill > 0) {
%            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
%            # add those boxes to the inactive list
%        }
%        ninact <- length(inactive)
%        kids <- list()  # subdivisions of active boxes
%        nkids <- 0
%        for (i in which(!kill)) {  # split active boxes into 4 parts
%            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
%            nkids <- nkids + 4
%        }
%        active <- kids
%        nact <- length(active)
%
%        iter <- iter + 1
%        write(c("iteration", iter, "nact", nact, "ninact", ninact, "lowbound",
%            lowbound), file = "log.out", ncolumns = 8, append = TRUE)
%    }
%
%
%    tmp <- t(vapply(X = c(active, inactive), FUN = function(box) {
%        c(box$lims.sigsqs, box$lims.sigsqe, box$bounds)
%    }, FUN.VALUE = c(sigsqs.lo = 0.1, sigsqs.hi = 0.1, sigsqe.lo = 0.1, sigsqe.hi = 0.1,
%        rll.lower = 0.1, rll.upper = 0.1)))
%    end_time <- Sys.time()
%    write(paste(lognote, "runtime: ", round(end_time - start_time, digits = 3)),
%          file = "log.out", ncolumns = 1, append = TRUE)
%    return(data.frame(tmp))
%}

\paragraph{\textcompute{findf}: The main function}

\begin{verbatim}
#' Learn RLL or Log-posterior for two variance MM.
#'
#' Learns the shape of the objective function within epsilon.
#'
#' This is the primary function that implements the branch-bound-kill
#' algorithm described in "Approximately Exact Calculations for Linear
#' Mixed Models" by Lavine & Hodges (2015). This an interative algorithm
#' that can require substantial computation time. It is recommended that
#' the user start with a conservative \code{maxit} and do additional
#' computations as necessary.
#'
#' The arguments of this function include five settings to control when
#' to stop branching (subdividing) a box:
#' \enumerate{
#'   \item when the upper and lower bounds are within \code{eps} of one another.
#'   \item when the width of \eqn{\sigma^2_e} is less than \code{delE}.
#'   \item when the width of \eqn{\sigma^2_s} is less than \code{delS}.
#'   \item when the upper bound is more than \code{M} below the highest global
#'   lower bound.
#'   \item when the number of iterations reaches \code{maxit}.
#'   }
#'
#' @param lines a dataframe that contains the constants that define the
#'   line represented by each term in the sum. Created as output from
#'   \code{\link{findlines}}.
#' @param startbox a list of boxes and their bounds, possibly the output
#'   from a previous call to \code{findf}. If left empty, will create
#'   a new startbox from a call to \code{makebox}.
#' @param eps a non-negative numeric indicating the tolerance within which you
#'   would like to learn the function. Default is 0.
#' @param delE a non-negative numeric indicating the width of \eqn{\sigma^2_e}
#'   beyond which the algorithm will stop branching boxes. Default is 0.
#' @param delS a non-negative numeric indicating the width of \eqn{\sigma^2_s}
#'   beyond which the algorithm will stop branching boxes. Default is 0.
#' @param M a non-negative numeric indicating the size of the buffer between
#'   the highest global lower bound and the upper bound of a given box past
#'   that box will now be branched further.  Default is \code{Inf}.
#' @param maxit a positive integer indicating the maximum number of iterations
#'   of the algorithm.  Default is 10.
#' @param ratio a logical indicating if \code{delE} and \code{delS} are specified as
#'   ZZQ.  Default is \code{FALSE}.
#' @param lognote a string to append to the "log.out" log file to annotate
#'   the purpose of each run of the algorithm. For use in benchmarking computation
#'   time. Default is \code{"summary"}.
#'
#' @return A list of boxes including their limits in \eqn{\sigma^2_e} and
#'   \eqn{\sigma^2_e} as well as their bounds. Running this function also results
#'   in the creation of a log file called "log.out" containing box counts at every
#'   iteration.

findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
                   ratio = FALSE, lognote = "summary") {

    start_time <- Sys.time()

    if (missing(lines)) {
        print("please supply lines")
        return
    }
    if (missing(startbox)) {
        startbox <- makebox(lines = lines, lims.sigsqs = c(0, max(lines$int.sigsqs[is.finite(lines$int.sigsqs)])),
            lims.sigsqe = c(0, max(lines$int.sigsqe[is.finite(lines$int.sigsqe)])),
            status = rep("straddle", nrow(lines)))
    }
    inactive <- list()
    ninact <- 0
    # check whether startbox is a box or a list of boxes
    active <- ifelse(length(startbox) == 4 && identical(names(startbox), c("lims.sigsqe",
        "lims.sigsqs", "status", "bounds")), list(startbox), startbox)
    nact <- length(active)

    lowbound <- -Inf
    iter <- 0

    # conditions under which a box becomes inactive
    killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
        # lb is a lower bound on max f; it changes at each iteration. M, eps,
        # delE, delS stay constant throughout the iterations.
        cond.low <- box$bounds[2] < lb - M
        cond.eps <- diff(box$bounds) < eps
        cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE, diff(box$lims.sigsqe) <
            delE)
        cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS, diff(box$lims.sigsqs) <
            delS)
        return(cond.low || cond.eps || cond.E || cond.S)
    }

    while (nact > 0 && iter < maxit) {
        # Find the lower bound of each box and the maximum of the lower bounds.  For
        # each active box, either make it inactive or divide it.
        low.act <- max(vapply(X = active, FUN = function(box) {
            box$bounds[1]
          }, FUN.VALUE = 0.1))
        lowbound <- max(lowbound, low.act)
        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
        nkill <- sum(kill)
        if (nkill > 0) {
            inactive[(ninact + 1):(ninact + nkill)] <- active[kill]
        }
        ninact <- length(inactive)
        kids <- list()
        nkids <- 0
        for (i in which(!kill)) {
            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)  # boxes are split into 4 parts
            nkids <- nkids + 4
        }
        active <- kids
        nact <- length(active)

@@

%\paragraph{\textcompute{killfunc}: determine whether to make a box inactive}
%\begin{verbatim}
%# conditions under which a box becomes inactive
%killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
%    # lb is a global (within startbox) lower bound on max(log(f));
%    #      it changes at each iteration.
%    #  M, eps, delE, delS stay constant throughout the iterations.
%    cond.low <- box$bounds[2] < lb - M
%    cond.eps <- diff(box$bounds) < eps
%    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
%                            diff(box$lims.sigsqe) < delE)
%    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
%                            diff(box$lims.sigsqs) < delS)
%    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
%}
%\end{verbatim}

\paragraph{\textcompute{makebox}: make a box}

\begin{verbatim}
#' Make a box with bounds
#'
#' Given the limits of the box, calculate its status relative to each line and
#' calculate the bounds on the objective function within that box.
#'
#' @param lims.sigsqe numeric vector of length 2 of the form \code{c(lower, upper)}
#'   containing the x-limits of the box.
#' @param lims.sigsqs numeric vector of length 2 of the form \code{c(lower, upper)}.
#'   containing the y-limits of the box.
#' @param status factor of length of \code{lines} denoting the location of the box
#'   relative to each line. If the box has a parent, these correspond to the status
#'   of that parent. The levels are \{above, below, straddle\}.
#' @param lines dataframe containing the constants that define the shape of each
#'   term in the sum. Output from \code{\link{findlines}}.
#'
#' @return A list containing:
#' \itemize{
#'   \item limits of the box in \eqn{\sigma^2_e} (numeric vector of length 2)
#'   \item limits of the box in \eqn{\sigma^2_s} (numeric vector of length 2)
#'   \item the status of the box relative to each line (factor of length of the
#'   number of terms in the sum)
#'   \item the bounds on the objective function (numeric vector of length 2 of
#'   the form \code{c(lower, upper)})
#'   }

makebox <- function(lims.sigsqe = NA, lims.sigsqs = NA, status = NA, lines) {

    # sanity checks
    if (missing(lims.sigsqs) || missing(lims.sigsqe))
        print("please supply lims.sigsqs and lims.sigsqe")
    if (missing(status))
        print("please supply status")
    if (missing(lines))
        print("please supply lines")

    # If the box has a parent then it inherits the parent's status.  But if the
    # parent straddles a line, we must check whether the child also straddles the
    # line.
    strad <- which(status == "straddle")
    status[strad] <- getstatus(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs,
        lines = lines[strad, ])

    # we could get some of the bounds from the parent, but it's just as easy to
    # recalculate them
    bounds <- getbounds(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs,
        status = status, lines = lines)

    return(list(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs, status = status,
        bounds = colSums(bounds)))
}
\end{verbatim}

\paragraph{\textcompute{splitbox}: split a box into 4 children}

\begin{verbatim}
#' Split a parent box into 4 children.
#'
#' Split a parent box into four equally sized boxes as part of the branching part
#' of the algorithm.
#'
#' @param box list containing the properties of a parent box. Created as output
#'   from \code{\link{makebox}}.
#' @param lines dataframe containing the constants that define the shape of each
#'   term in the sum. Output from \code{\link{findlines}}.
#'
#' @return a list of the four child boxes.

splitbox <- function(box, lines) {
    # split a box into four children
    NW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
        lines = lines))
    NE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
        lines = lines))
    SW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
        lines = lines))
    SE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
        lines = lines))
    return(list(NW, NE, SW, SE))
}
\end{verbatim}

\paragraph{\textcompute{getstatus}: determine where a box is, relative to the lines}

\begin{verbatim}
#' Get the location of a box relative to lines.
#'
#' Computes the location of a box with given limits relative to each of
#' the lines.
#'
#' @param lims.sigsqe numeric vector of length 2 of the form \code{c(lower, upper)}
#'   containing the x-limits of the box.
#' @param lims.sigsqs numeric vector of length 2 of the form \code{c(lower, upper)}
#'   containing the y-limits of the box.
#' @param lines dataframe containing the constants that define the shape of each
#'   term in the sum. Output from \code{\link{findlines}}.
#'
#' @return A factor of length of \code{lines} denoting the location of the box
#'   relative to each line.

getstatus <- function(lims.sigsqe, lims.sigsqs, lines) {
    # Is a box above, below, or straddling the lines with these slopes and
    # intercepts?

    # value of the lines at the left side of the box
    tmp1 <- lines$int.sigsqs + lines$slope * lims.sigsqe[1]
    tmp1[is.infinite(tmp1)] <- NA
